---
title: "lab_03"
author: "Derek Willis"
date: "2023-09-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## TASK: Load libraries

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse
library(tidyverse)

# Load janitor
library(janitor)
```

## TASK: Load data and clean up the column names

Load the two datasets we worked with in the pre_lab: the Maryland payments to vendors and Maryland grants and loans AND, using janitor, change the column names so that they are lower-cased and any spaces are replaced with underscores.

```{r}
# Load in the Maryland state agency payments data, making sure that it correctly sets the datatypes using all of the rows

maryland_payments <- read_csv ("data/State_of_Maryland_Payments_Data__FY2008_to_FY2024.csv", guess_max=373564)

cleaned_maryland_payments <- maryland_payments |>
  clean_names()

# Load in the Maryland grants and loan data

maryland_grants_and_loans <- read_csv ("data/State_of_Maryland_Grant_and_Loan_Data__FY2009_to_FY2022.csv")

cleaned_maryland_grants_and_loans <- maryland_grants_and_loans |>
  clean_names() |> 
  rename(source = grantor) |> 
  mutate(source = str_to_upper(source), grantee = str_to_upper(grantee), description = str_to_upper(description), category = str_to_upper(category))
```

## Answer questions

Q1: You want to know how much money state agencies have paid to the Prince George's County School system, but to do that you'll need to find all of the variations of how the school system might be listed in the payments data. Look at & search the records in the dataframe to start your search, starting with a broad search and then narrowing it down. Then write code to find rows where the vendor name contains some version of the county's school system. 

HINT: you should use filter() just once, and inside your filter you want to look for the variations one at a time. Think "this version OR that version."

When you have found as many as possible, save the result to a new dataframe called `pg_schools`. How many records did you find? 

A1:I found 197 records for pg_schools and state agencies paid $16,717,094,414.


```{r}
pg_county_distinct_vendors <- distinct(cleaned_maryland_payments, vendor_name)

pg_schools <- cleaned_maryland_payments |> 
  filter(str_detect(vendor_name, "BOARD OF EDUCATION OF PRINCE GEORGE") | 
           str_detect(vendor_name, "BOARD OF EDUCATION OF PRINCE GEORGE'S COUNTY") |
           str_detect(vendor_name, "PRINCE GEORGE'S COMM COLLG") |
           str_detect(vendor_name, "PRINCE GEORGE'S COMMUNITY COLLEGE") |
           str_detect(vendor_name, "PRINCE GEORGES COMMMUNITY COLLEGE FOUNDATION, INC") |
           str_detect(vendor_name, "PRINCE GEORGES COMMUNITY COLLEGE") |
           str_detect(vendor_name, "PRINCE GEORGES COUNTY PUBLIC SCHOOL") |
           str_detect(vendor_name, "PRINCE GEORGES COUNTY PUBLIC SCHOOLS") |
           str_detect(vendor_name, "PRINCE GEORGES COUNTY SCHOOL") |
           str_detect(vendor_name, "PRINCE GEORGES PUBLIC SCHOOLS"))

pg_school_total <- pg_schools |> 
  summarise(
    total = sum(amount)
  )

  
```

Q2: Using that `pg_schools` dataframe, write code to calculate the total amount of payments to the Prince George's County School system for each fiscal year. Which fiscal year had the largest total amount? Are there any aspects of the results that stand out to you?

A2: 2022 had the largest total amount $1,623,881,847. An aspect that struck me was that this year (2023), the PG county school system received 1,460,851,630 less than last year. 

```{r}

fiscal_year_pg_schools <- pg_schools |> 
  group_by(fiscal_year) |> 
  summarise(
    total = sum(amount)
  ) |> 
  arrange(desc(fiscal_year))
```

Q3. You're interested in examining the vendors paid for capital projects at the University of Maryland. Search the payments data to find those payments, then write code to produce a list of each vendor and the total amount it received. 

Look carefully at that list - what cleaning will it require to ensure that you have only one row per vendor and zip code? Provide at least three examples of vendor names that are similar and should be combined into one. Describe in words how you might clean that data so that any nearly identical vendors would look identical - what would you change and how?

A3. I would clean the data with the distinct function, ensuring that there is only one row per vendor and zip code. Three examples of vendor names that are similar: A MORTON THOMAS & ASSOC INC and A MORTON THOMAS & ASSOCIATES INC, A&I  INC and A&I INC., AEROSOL MONITORING & ANALYSIS INC and AERSOL MONITORING & ANALYSIS INC. According to CHAT.GPT (I asked it how would I clean the data so that nearly identical data values would look identical in R?) and datablist.com, I would "use functions like duplicated() or distinct() from the dplyr package to identify exact duplicates or similar records" and use the the Levenshtein distance to get a similarity score through the stringdistmatrix() to get identical results. The Levenshtein distance "quantifies distance or similarity between two strings by calculating the minimum number if single-character edits required to transform one string into another."


```{r}
capital_projects_UMD <- cleaned_maryland_payments |> 
  filter(str_detect(agency_name, "CAPITAL PROJECTS- UNIVERSITY OF MARYLAND")) 

capital_projects_UMD_vendors <- capital_projects_UMD |> 
  group_by(vendor_name) |> 
  summarise(
    total = sum(amount)
  )

```

Q4: You want to understand how much state grant money went to stem cell research projects each fiscal year. Search the data to find grants that fit that description. Once you find which column(s) contain the information you want, your code should find just the grants for stem cell research in a single filter. Then calculate the total amount of money for each fiscal year. Write a sentence summarizing how stem cell research funding has changed over time.

A4: Stem cell research had the most funding when it began in 2009, then had the greatest amount since in 2012 and dipped to its lowest level in 2021 before increasing again in 2022. 

```{r}
stem_cell_research <- cleaned_maryland_grants_and_loans |> 
  group_by(fiscal_year) |> 
  filter(str_detect(description, "STEM CELL RESEARCH GRANT")) |> 
  summarise(
    total = sum(amount)
  )

```

Q5: Using the Maryland grants and loan data, find records where the zip code is the one where the University of Maryland-College Park is located in and produce a list of grantees from that zip code, along with the total amount they received. Write a paragraph explaining how many grantees are there, what you would need to do to clean the results so that you had one record per grantee and which grantees are most interesting to you?

A5: There are 18 per this output but I would need to clean the results by including a case_when() that would group grantees that are the University of Maryland College Park (in College Park) in the same entry rather than have 3 different entries for the same thing ("UNIVERSITY OF MARYLAND," "UNIVERSITY OF MARYLAND COLLEGE PARK," "UNIVERSITY OF MD,COLLEGE PARK"). The one that is the most interesting to me is the UMD Office of Research Administration and Advancement because it receives the second most amount of funding. 

```{r}
#20742

zip_code_umd_grantees <- cleaned_maryland_grants_and_loans |> 
  group_by(grantee) |> 
  filter(str_detect(zip_code, "20742")) |> 
  summarise(
    total = sum(amount)
  ) |> 
  arrange(desc(total))

```
