---
title: "lab_06"
author: "derek willis"
date: "2022-03-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# load your libraries
library(tidyverse)
library(lubridate)
library(janitor)
library(dplyr)
```

## Getting Started

This week we'll be combining and joining data to make it more useful and to ask some questions about it. We'll use some of our 911 overdose call data to do this. The first thing we want to do is to combine multiple counties' data into a single dataframe so that we can ask some questions. First, let's combine data from Cecil, Carroll and Allegany counties into a new dataframe.

*Before* you combine them you'll need to clean up some of the column names to make the data easier to work with - make sure you have loaded the library to do that. You also need to make sure that each column has the same name and datatype (you can check the datatypes using `glimpse`). If any of them does not, you need to fix that.

```{r}
# load and combine the call data from those three counties

cecil_911 <- read_csv("data/cecil_911.csv")
carroll_911 <- read_csv("data/carroll_911.csv")
allegany_911.csv <- read_csv("data/allegany_911.csv")
glimpse(carroll_911_clean)

carroll_911_clean <- carroll_911 |> 
  clean_names()

combined_county_911 <- bind_rows(list(allegany_911.csv, carroll_911_clean, cecil_911))
View(combined_county_911)
```

Then we'll join some demographic data with Baltimore City 911 calls and ask some questions of that. For that, you'll need to load the Baltimore City 911 data and the ZCTA demographic data, and join them based on the zip code. Remember, the zip codes in both dataframes must be the same datatype (and should be characters). You should start with your Baltimore 911 dataframe in creating your join. 

```{r}
# load and join the Baltimore-related data

baltimore_911 <- read_csv("data/baltimore_911.csv")
maryland_zcta <- read_csv("data/maryland_zcta.csv")
maryland_zcta_clean <- maryland_zcta |> 
  clean_names() |> 
  rename(zip_code = zcta5n)
  
baltimore_maryland_combined_911 <- baltimore_911|> left_join(maryland_zcta_clean)

```

## Answer questions

Q1. Let's start with our data from Allegany, Cecil and Carroll counties. Write code to generate a dataframe that shows the total number of calls for each county. What's the order of counties from most calls to least?

A1. The county with the most calls was Carroll County with 448, then Cecil with 438, and then Allegany with 418. 

```{r}
most_calls_county <- combined_county_911 |> 
  group_by(county) |> 
  summarise(
    count = n()) |> 
  arrange(desc(count))
  head(most_calls_county)
```

Q2. In your combined dataframe that you used in A1, add a column for the month of each call, then write code to show the total number of calls per county and month. Are there any outliers or noteworthy results? Describe the general pattern of the results. 

Then calculate the most calls per county and location. Which location had the most calls - where and what is that location?

A2.I asked Chat. GPT how "What code do I use in R to add a column with just the month when I have a date column that goes year, month, date?" and it gave me this code: df$month_column <- month(df$date_column). The most calls were in Cecil and Allegany in the month of August and Allegany had the most calls after Cecil in the months of August, September, April, July, and October. The locdation with the most calls was in Cecil County at 314 GROVE NECK RD EARLEVILLE, MD with 15 calls.

```{r}
combined_county_911 |> 
  mutate(month = month(date)) |> 
  group_by(county, month) |> 
  summarise(
    count = n()) |> 
  arrange(desc(count))

combined_county_911 |> 
  group_by(county, location) |> 
  summarise(
    count = n()) |> 
  arrange(desc(count))
  
```

Q3. Let's turn to the Baltimore City data. You're interested in exploring calls in zip codes with a high percentage of children under 18. Write code to isolate calls in zip codes where the percentage of under 18 population is at least 75% and then show the zip code and population and how many calls occurred in each of those zip codes. 

Where are those zip codes in the city?

Add a column to your code that calculates the number of calls per 1,000 people for those zip codes. To calculate a per capita rate, you might find this short guide handy: https://observablehq.com/@palewire/per-capita-calculator.

Which zip code has the highest rate of calls per 1,000 people? Find the neighborhoods in that zip code that are listed in the data - you can use group_by or distinct to do this. What are some of those neighborhoods, and what else can you tell me about the population there?

A3. The zip codes that had at least a 75% of under 18 population were 21201, 21202, 21205, 21213, 21216, 21217, 21223. The zip code that had the highest percentagte of children under 18 per 1,000 people was 21217. Some of the neighborhoods that are in zip code 21217 are Penn North, Sandtown-Winchester, Upton, Druid Heights, and Parkview/Woodbrook. In that zip code, there is a population of 37111 and the demographic with the highest percentage of that area is Black with 70% of the population renting.

```{r}
baltimore_maryland_combined_911 |> 
   group_by(zip_code, pct_under_18, population) |>
   filter(pct_under_18 >= 75) |> 
   summarise(
    count = n()) |> 
  mutate(per_capita = (count/population)*1000) |> 
  arrange(desc(per_capita))

baltimore_maryland_combined_911 |> 
  group_by(zip_code, neighborhood) |> 
  filter(zip_code == "21217") |> 
  summarise(
    count = n()) |> 
  arrange(desc(count))

baltimore_maryland_combined_911 |> 
   group_by(zip_code, pct_under_18, population, neighborhood) |>
   filter(pct_under_18 >= 75, zip_code == "21217") |> 
   summarise(
    count = n()) |> 
  mutate(per_capita = (count/population)*1000) |> 
  arrange(desc(per_capita))
  



```

Q4. Choose your own adventure: I want you to come up with a question you could ask of the Baltimore City data that combines both the calls and the demographic information. You must use both demographics and the overdose call data in your work.

A4: What is the neighborhood with the highest percentage of Hispanic people and the least? What is the percentage per capita of 10,000 people for 911 calls for both neighborhoods? 

**Many neighborhoods have the highest Hispanic percentage as being 18.5% but the neighborhood with the most number of calls and highest Hispanic percentage is Baltimore Highlands and the neighborhood with the lowest percentage of Hispanic people is Morgan State University. The percent of 911 calls per 10,000 people in Baltimore Highlands is 7.33% however it is 64.24% in Morgan State University in the 21251 zip code but 0.60% in the Morgan State University University in the 21218 zip code. **

```{r}

baltimore_maryland_combined_911 |> 
  group_by(pct_hispanic, neighborhood) |> 
  summarise(
    count = n()
  ) |> 
  arrange(desc(count))


most_hispanic_neighborhood <- baltimore_maryland_combined_911 |> 
  group_by(population, neighborhood) |>
   summarise(
    total = n()) |> 
  filter(neighborhood == "Baltimore Highlands") |> 
  mutate(per_capita = (total/population)*10000) |> 
  arrange(desc(per_capita))

leas_hispanic_neighborhood <- baltimore_maryland_combined_911 |> 
  group_by(population, neighborhood, zip_code) |>
   summarise(
    total = n()) |> 
  filter(neighborhood == "Morgan State University") |> 
  mutate(per_capita = (total/population)*10000) |> 
  arrange(desc(per_capita))



```
