---
title: "lab_10"
author: "Derek Willis"
date: "2023-04-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

* Our usual libraries, plus some of the map-specific ones from the pre_lab. You'll also be fetching Census data in Q3. Read the questions carefully!

## Load libraries and establish settings

**Task** Create a codeblock and load appropriate packages and settings for this lab.

```{r}
# Load libraries
library(tidyverse)
library(sf)
library(janitor)
library(tigris)
install.packages("mapview")
library(mapview)
library(tidycensus)
census_api_key()
```

Let's explore Maryland elections, power outages and electric vehicle data and make some maps! 

## Questions 

**Q1.** Maryland voters made history in November by electing the state's first two African-American statewide office holders: Wes Moore as governor and Anthony Brown as attorney general. You want to compare the two's electoral performances. Using the `md_statewide_county.csv` file in the data folder, calculate the total number of votes for both the governor's race and the attorney general's race, then percentages for both Moore and Brown _based on the total of their respective contests_. Then calculate the difference between their percentages.

```{r}
md_statewide_county <- read_csv("data/md_statewide_county.csv") |> 
  clean_names()

total_votes_moore_and_brown <- md_statewide_county |> 
  mutate(total_gov = cox + moore + lashar + wallace + harding + sefcik + gov_writein,
         total_ag = peroutka + brown + ag_writein,
         percent_moore = (moore/total_gov)*100,
         percent_brown = (brown/total_ag)*100,
         difference = (percent_moore - percent_brown),
         geoid = as.character(geoid)) |> 
  arrange(desc(difference))

glimpse(total_votes_moore_and_brown)

```

Get the geometry for Maryland's counties using tigris and, joining that dataframe to your results dataframe, make a county-level map of the differences between Moore and Brown's percentages. What is the story here, and what is the county with the most interesting results?

**A1. The counties that had the biggest difference in voting percentages between Moore and Brown were Queen Anne's, Kent and Carroll counties. The county with the most interesting results was Queen Anne's County because Moore got more than 7% more votes for Governor and than Brown for Attorney General were very big for somewhere very close to the state capital.** 

```{r}
md_counties <- counties |>
  filter(STATEFP == "24") |> 
  clean_names()
glimpse(md_counties)

md_gov_ag_race_counties <- md_counties |> left_join(total_votes_moore_and_brown, join_by(geoid)) |> 
  gli

ggplot() +
  geom_sf(data=md_gov_ag_race_counties, aes(fill=difference)) +
  scale_colour_viridis_b(option="magma") +
  theme_minimal() #+

ggplot() +
  geom_sf(data=md_gov_ag_race_counties, aes(fill=difference)) +
  theme_minimal() +
  scale_fill_viridis_b(option="magma")

mapview(md_gov_ag_race_counties, zcol = 'difference')
```

**Q2.** Make a county-level map to explore the number of customers impacted by power outages in Maryland. Load the `county_outages_march2023.csv` file in the data folder and calculate a rate of customers with outages per 10,000 total customers for each county. Then join that dataframe to your Maryland counties dataframe you made in Q1 and create a county-level map of the results; you may need to tweak the fill scale to make the distinctions clearer. What's the potential story here? Given access to other months' outage data, what would your next steps be in reporting a story?

**A2.Talbot County has by far the most rate of outages per 10,000 customers. With 1,989.84 customers with outages per 10,000 customers. The potential story is why does Talbot County have such a high rate of outages while the surrounding areas do not. ** 

```{r}
county_outages_march2023 <- read_csv("data/county_outages_march2023.csv")

rate_of_customers <- county_outages_march2023 |> 
  mutate(rate = (outage_customers/total_customers)*10000) |>
  arrange(desc(rate))

rate_in_counties <- md_counties |> left_join(rate_of_customers, by=c('name'='county')) |>
  glimpse()

mapview(rate_in_counties, zcol = 'rate')

```

**Q3.** You've been assigned a story about the use of electric vehicles in Maryland and you want to make a map of the prevalence of electric vehicles in each county. Load the `ev_2020_2023.csv` file, which has the number of electric and hybrid vehicles per county from July 2020 and March 2023 and calculate a percentage change for each county (remember, that's `new-old/old`). Using the Maryland counties dataframe you made for Q1, make a county map showing the percentage change.What counties stand out to you? What questions do you have about the changes you're seeing?

Next, use tidycensus to retrieve county population data for Maryland from 2021 (you did this in the pre_lab) and join it to your vehicles dataframe. HINT: you may not need to join it on the county name, and you should already have the geometry!

Then, calculate a per capita rate of electric & hybrid vehicles using the March 2023 figure and the population estimate. You will need to choose the "per capita" value.

Finally, make a second county map showing the per capita rate. What counties stand out on this map? How is it different from the first one you made?

**A3.Caroline County has had the highest percent change in increase of electric and hybrid cars with an increase of over 400%. Electric and hybrid cars have seen a drastic increase all over Maryland with even the county with the fewest electric cars (Dorchester with 50 in 2020) with an over 100% increase. What has made these cars so much more accessible and affordable that now a lot more people have them?**

**Now in the map it shows that Montgomery and Howard Counties have the highest number of electric and hybrid cars per capita. This map differs from the first one because they highlight different counties. The first one highlights Queen Anne's, Kent, Carroll, Prince George's counties and Baltimore City.**

```{r}
ev_2020_2023 <- read_csv("data/ev_2020_2023.csv")

electric_cars_change <- ev_2020_2023 |> 
  mutate(perc_change = ((march_2023-july_2020)/july_2020)*100)

electric_cars_md <- md_counties |> left_join(electric_cars_change, by=c('name'='county')) |> 
  arrange(desc(perc_change))


md_county_population <- get_acs(geography = "county",
              variables = c(population = "B01001_001"),
              year = 2021,
              state = "MD") |> 
  clean_names()

electric_cars_pop_md <- electric_cars_md |> left_join(md_county_population, join_by(geoid)) |> 
  rename(county = name.x, name = name.y) |> 
  mutate(total = july_2020 + march_2023) |> 
  mutate(per_capita = (total/estimate)*1000) |> 
  arrange(desc(per_capita)) |> 
  #intptlat = as.numeric(intptlat) |> 
  #intptlon = as.numeric(intptlon) |> 
  glimpse()

st_as_sf(electric_cars_pop_md, coords = c("geometry"))
mapview(electric_cars_pop_md, zcol = 'per_capita')

electric_cars_pop_md <- st_as_sf(electric_cars_pop_md, coords = c("intptlat", "intptlon"), crs = 4326)


```

